The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-19 script/en/chapters/numbertheory.tex
CT script: Minor corrections, mostly layout, de + en; final now for CT 1.4.30 Beta04

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem~\ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter~\ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\noindent The following three examples\footnote{%
\noindent See \hyperlink{AppArith1}{Appendix E of this chapter} for the source
code to compute the tables using Sage.}
illustrate the properties of multiplicative inverses.

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
We have\footnote{%
See \hyperlink{AppArith2}{Appendix E of this chapter} for source code
implementing the square and multiply method in Sage, which can be used
to reproduce the calculations above.}:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}
\end{example}

@@

%\newpage
\pagebreak

\begin{center}
\fbox{\parbox{15cm}{{\em Eric Berne\footnotemark:\index{Berne, Eric}}\newline
Mathematical game theory postulates players who respond rationally.
Transactional game theory, on the other hand, deals with games that are 
not rational, perhaps even {\bf irrational and thereby closer to reality}.}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{Eric Berne, ``Games People Play'', 
     rororo, (c) 1964, page 235.}
\vskip +4pt

@@

\begin{example}{:}
\begin{itemize}
\item  $n=70=2*5*7 \Longrightarrow $ using theorem~\ref{J_of_p1..pk}: $ J(n)= 1\cdot 4 \cdot 6 =24.$
\item  $n=9=3^2 \Longrightarrow$ using theorem~\ref{J_of_n}: $ J(n)= 3^1\cdot 2 =6,$ because  $\mathbb{Z}_9^* =\{ 1,2,4,5,7,8\}.$
\item $n = 2,701,125 = 3^2 * 5^3 * 7^4 \Longrightarrow $ using theorem~\ref{J_of_n}: 
$$J(n) = [3^1 * 2] * [5^2 * 4] * [7^3 * 6] = 1,234,800.$$
\end{itemize}
\end{example}

@@

\begin{example}{ 1:}\\
The following table~\ref{expmod11}\footnote{%
See \hyperlink{AppArith3a1}{Appendix E of this chapter} for the source
code to generate the table using Sage.} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$/
\end{example}

@@

\begin{example}{ 2:}\\
The following table~\ref{expmod45}\footnote{%
See \hyperlink{AppArith3b}{Appendix E of this chapter} for the source
code to generate the table using Sage.} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$.
\end{example}

@@

\noindent The requirement/condition $gcd(7,45)=1$ is fulfilled.
The table 'values of $a^{i}$ mod $45$' shows that the number $7$ is not a primitive root of $45$, because $ord_{45}(7) = 12 \not= 24 = J(45)$.
\end{example}

\begin{example}{ 4:}\\
The following table~\ref{modexp46}\footnote{%
See \hyperlink{AppArith3c}{Appendix E of this chapter} for the source
code to generate the table using Sage.} answers the question as to
whether the number $7$ is a primitive root of $46$. 
The requirement/condition $gcd(7,46)=1$ is fulfilled.
\end{example}

\begin{table}[ht]
{\textmd \small
\begin{center}  
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & \textbf{1}  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}
\caption{Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$}
\label{SrcArith3c}\label{modexp46}
\end{table}

$J(46)$ is calculated using theorem~\ref{thm-zth-phipq}: $J(46) = J(2*23) = 1*22 = 22$.
The number $7$ is a primitive root of $46$,  because $ord_{46}(7) = 2 = J(46)$.

\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{modexp46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  } Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}
The multiplicative group $Z_n^*$ contains all values from $1$ to $n-1$.

@@

Further information concerning \hyperlink{AppArith3a2}{primitive roots\index{Primitive root}}
can be found in appendix E of this chapter. %\ref{primitive-roots-with-sage}. xxxxxprro
% appendix E eigentlichnicht gut, aber~\ref{primitive-roots-with-sage}. liefert 4.13.5!  xxxxxprro

} is extremely difficult to solve.
\end{remark}

@@

\label{SrcArith4a}
\noindent All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code
to do RSA encryption using Sage. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

\noindent We will work through the example in 2 variants. The steps 1 to 3 are common for both.
\par
\noindent {\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix E of this chapter} for the source
code to factorize the number $J(n)$ using Sage.
Using CrypTool\index{CrypTool} you can solve this with the 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorization of a Number}.}. \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ (should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorized but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\label{SrcArith4c}
\noindent Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix E of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 198967 051405 254571 115318 014251}

@@

The order $\text{ord}_m(a)$ of a number $a$ in the multiplicative
group $\mathbf{Z}_m^{\ast}$ is the smallest number $i \geq 1$ such that
$a^i \equiv 1 \pmod{m}$ holds (see chapter ``\nameref{MultOrdPrimitveRoot}'').

To create table~\ref{expmod11} on page~\pageref{SrcArith3a} we can print
all exponentiation $a^i \pmod{11}$ as follows:

@@

Computing a primitive root (see chapter ``\nameref{MultOrdPrimitveRoot}'')
in Sage is very straightforward. If \verb!n! is an integer, the command
\verb!primitive_root(n)! computes a primitive root of the multiplicative group
$(\mathbf{Z} / n \mathbf{Z})^{\ast}$, if one exists.
Where $n$ is prime, then this is the same as calculating a primitive root of
$\mathbf{Z} / n \mathbf{Z}$.

>>>>

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem~\ref{thm-zth-ordp} in \hyperlink{Chapter_ElementaryNT_9}
{chapter~\ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\noindent The following three examples\footnote{%
\noindent See \hyperlink{AppArith1}{Appendix of this chapter} for the source
code to compute the tables using Sage.}
illustrate the properties of multiplicative inverses.

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
We have\footnote{%
See \hyperlink{AppArith2}{Appendix of this chapter} for source code
implementing the square and multiply method in Sage, which can be used
to reproduce the calculations above.}:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}
\end{example}

@@

% --------------------------------------------------------------------------
\pagebreak
\begin{center}
\fbox{\parbox{15cm}{{\em Eric Berne\footnotemark:\index{Berne, Eric}}\newline
Mathematical game theory postulates players who respond rationally.
Transactional game theory, on the other hand, deals with games that are 
not rational, perhaps even {\bf irrational and thereby closer to reality}.}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{Eric Berne, ``Games People Play'', 
     rororo, (c) 1964, page 235.}
\vskip +4pt

@@

\pagebreak  % Oder \newpage oder \vskip +40pt 
\begin{example}{:}
% \nopagebreak dazwischen nützte nichts --> vskip !!
\begin{itemize}
%\nopagebreak Nützte nichts !!
\item  $n=70=2*5*7 \Longrightarrow $ using theorem~\ref{J_of_p1..pk}: $ J(n)= 1\cdot 4 \cdot 6 =24.$
\item  $n=9=3^2 \Longrightarrow$ using theorem~\ref{J_of_n}: $ J(n)= 3^1\cdot 2 =6,$ because  $\mathbb{Z}_9^* =\{ 1,2,4,5,7,8\}.$
\item $n = 2,701,125 = 3^2 * 5^3 * 7^4 \Longrightarrow $ using theorem~\ref{J_of_n}: 
$$J(n) = [3^1 * 2] * [5^2 * 4] * [7^3 * 6] = 1,234,800.$$
\end{itemize}
\end{example}

@@

\begin{example}{ 1:}\\
The following table~\ref{expmod11}\footnote{%
See \hyperlink{AppArith3a1}{Appendix of this chapter} for the source
code to generate the table using Sage.} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$/
\end{example}

@@

\begin{example}{ 2:}\\
The following table~\ref{expmod45}\footnote{%
See \hyperlink{AppArith3b}{Appendix of this chapter} for the source
code to generate the table using Sage.} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$.
\end{example}

@@

\noindent The necessary, but not sufficient requirement/condition $gcd(7,45)=1$ is fulfilled.
Table~\ref{expmod45} shows that the number $a=7$ is not a primitive root of $45$, because $ord_{45}(7) = 12 \not= 24 = J(45)$.
\end{example}

\begin{example}{ 4:}\\
The following table~\ref{modexp46}\footnote{%
See \hyperlink{AppArith3c}{Appendix of this chapter} for the source
code to generate the table using Sage.} answers the question as to
whether the number $a=7$ is a primitive root of $46$. 
The  necessary, but not sufficient requirement/condition $gcd(7,46)=1$ is fulfilled.\\
$J(46)$ is calculated using theorem~\ref{thm-zth-phipq}: $J(46) = J(2*23) = 1*22 = 22$.
The number $7$ is a primitive root of $46$,  because $ord_{46}(7) = 2 = J(46)$.
\end{example}

\begin{table}[ht]
{\textmd \small
\begin{center}  
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & \textbf{1}  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}
\caption{Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$}
\label{SrcArith3c}
\label{modexp46}
\end{table}

\vskip +10 pt
\begin{theorem}\label{thm-zth-ordp}\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{modexp46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  } Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.
\end{theorem}
The multiplicative group $Z_n^*$ contains all values from $1$ to $n-1$.

@@

Further information concerning \hyperlink{AppArith3a2}{primitive roots\index{Primitive root}}
can be found in the appendix of this chapter. %\ref{primitive-roots-with-sage}. xxxxxprro
% appendix E eigentlich nicht gut, aber~\ref{primitive-roots-with-sage}. liefert 4.13.5!  xxxxxprro
} is extremely difficult to solve.
\end{remark}

@@

\label{SrcArith4a}
\noindent All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix of this chapter} for source code
to do RSA encryption using Sage. \\
You can also encrypt the message with CrypTool\index{CrypTool} via the menu path 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} RSA Demonstration}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

\noindent We will work through the example in 2 variants. The steps 1 to 3 are common for both.
\par
\noindent {\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503,~q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix of this chapter} for the source
code to factorize the number $J(n)$ using Sage.
Using CrypTool\index{CrypTool} you can solve this with the 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorization of a Number}.}. \\
{\bf 2.} $e=65,537$ \\
\strut\quad\ (should\footnote{%
                See footnote~\ref{foot:Selection-of-e} on page
                \pageref{foot:Selection-of-e}.}
                lie between $509$ and $255,015$ and must\footnote{%
       $e$ cannot, therefore, be $2, 127$ or $251$
       ($65,537 = 2^{16}+1$) ($255,016 = 2^{3}*127*251$).\\
       In real life, $J(n)$ is not factorized but rather the Euclidean
       algorithm is used for the selected e to guarantee that
       ${\rm gcd}(e,J(n))=1$.
                 } be relatively prime to $255,016$).\\
{\bf 3.} $d=231,953$ \\
\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}255,016)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\label{SrcArith4c}
\noindent Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
\noindent Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix of this chapter} for the source
code for RSA exponentiation using Sage.
}: \\
{\tt 198967 051405 254571 115318 014251}

@@

The order $\text{ord}_m(a)$ of a number $a$ in the multiplicative
group $\mathbf{Z}_m^{\ast}$ is the smallest number $i \geq 1$ such that
$a^i \equiv 1 \pmod{m}$ holds
(see chapter~\ref{MultOrdPrimitveRoot}, ``\nameref{MultOrdPrimitveRoot}'').
To create table~\ref{expmod11} on page~\pageref{SrcArith3a} we can print
all exponentiation $a^i \pmod{11}$ as follows:

@@

Computing a primitive root (see chapter~\ref{MultOrdPrimitveRoot},
``\nameref{MultOrdPrimitveRoot}'')
in Sage is very straightforward. If \verb!n! is an integer, the command
\verb!primitive_root(n)! computes a primitive root of the multiplicative group
$(\mathbf{Z} / n \mathbf{Z})^{\ast}$, if one exists.
Where $n$ is prime, then this is the same as calculating a primitive root of
$\mathbf{Z} / n \mathbf{Z}$.
